---
title: "Section4: Population Scale Analysis"
author: "Ellice Wang (PID: A16882742)"
date: February 13, 2025
toc: True
format: pdf
---

## Q13: read expression data R
> determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes

The median expression levels for each genotype are: 
A/A: 31.25
A/G: 25.065
G/G: 20.074

```{r}
# import dplyr and ggplot
library(dplyr)
library(ggplot2)
```


```{r load table}
url <- "https://bioboot.github.io/bimm143_W25/class-material/rs8067378_ENSG00000172057.6.txt"

exp_data <- read.table(url)
head(exp_data)
```

```{r}
unique(exp_data$geno)

A_G <- exp_data[exp_data$geno == "A/G",]
summary(A_G)

A_A <- exp_data[exp_data$geno == "A/A",]
summary(A_A)

G_G <- exp_data[exp_data$geno == "G/G",]
summary(G_G)

```

## Q14: boxplot
> Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

```{r}
ggplot(exp_data, aes(x=geno, y=exp, fill = geno)) + 
  geom_boxplot(outliers = FALSE, notch = TRUE) + 
  geom_jitter(color="black", position=position_jitter(0.2), alpha=0.3)
```


