---
title: "Class 10: Halloween Mini-Project"
author: "Ellice Wang"
date: February 4, 2025
toc: true
format: pdf
prefer-html: true
---

Today we will examine data from 538 on common Halloween candy. We will use ggplot, dplyr, and PCA to make sense of this multivariate dataset. 

```{r}
# load in libraries
library(ggplot2)
library(dplyr)
```


# 1. Importing candy data

```{r}
# by url
url <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"

candy <- read.delim(url, sep = ",")
```


```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)
```
> Q1. How many different candy types are in this dataset?

There are 85 different candies in this dataset. 

```{r}
nrow(candy)
```

> Q2. How many fruity candy types are in the dataset?

There are 38 fruity candy types in this dataset.

```{r}
sum(candy$fruity)
```

# 2. What is your favorite candy?

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

My favorite candy is Reese's Peanut Butter cups. It has a winpercent value of **84.18029**.

```{r}
candy["Reese's Peanut Butter cup",]$winpercent
```

> Q4. What is the winpercent value for “Kit Kat”?

The win percent value for Kit kat is **76.7686**. 

```{r}
candy["Kit Kat","winpercent"]
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

The win percent value for Tootsie Roll Snack Bars is **49.6535**. 


```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```

```{r}
library("skimr")
skim(candy)
```
> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

It seems that winpercent row is different than the majority of the other columns in the dataset. The other columns range from 0 to 1 while winpercent goes from 14%-84%. 

> Q7. What do you think a zero and one represent for the candy$chocolate column?

It represents whether or not the candy contains chocolate (1) or if it does not (0). 

```{r}
head(candy)
```

> Q8. Plot a histogram of winpercent values

```{r}
ggplot(candy, aes(x=winpercent)) + 
  geom_histogram(bins=13, fill="dodgerblue") 
```

> Q9. Is the distribution of winpercent values symmetrical?

No the distribution is not symmetrical. The data is slightly skewed towards the left side of the graph, or a lower win percent value. 

> Q10. Is the center of the distribution above or below 50%? 

The center of the distribution is around 50%. The median is below 50% but the mean is at 50.32%. 

```{r}
summary(candy$winpercent)
```


> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

On average, the choclate candy is ranked higher than the fruit candy. 

```{r}

chocolate_win <- candy$winpercent[(candy$chocolate) == 1]
mean.choc_win <- mean(chocolate_win)
fruit_win <- candy$winpercent[as.logical(candy$fruity)]
mean.fruit_win <- mean(fruit_win)

paste("chocolate:", mean.choc_win, "fruit:", mean.fruit_win, sep=" ")
```

> Q12. Is this difference statistically significant?

This different is statistically significant. The p-value of the chocolate and fruit data is < 0.05 which suggests that the difference is statistically significant. 

```{r}
t.test(chocolate_win, fruit_win)
```
# 3. Overall Candy Rankings
> Q13. What are the five least liked candy types in this set?

The five least liked candy types are: Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters. 

```{r}
head(arrange(candy, winpercent), 5)
```

> Q14. What are the top 5 all time favorite candy types out of this set? 

The top 5 all time favorite candy types is Reese's Peanut Butter cup, Reese's Miniatures, Twix, Kit Kat, and Snickers. 

```{r}
head(arrange(candy, desc(winpercent)), 5)
```

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col(position = "dodge")
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col()
```

```{r}
my_cols=rep("darkgrey", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate4"
my_cols[as.logical(candy$bar)] = "sandybrown"
my_cols[as.logical(candy$fruity)] = "violetred1"
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

> Q17. What is the worst ranked chocolate candy?

The worst ranked chocolate candy are Sixlets. 

> Q18. What is the best ranked fruity candy?

The best ranked fruit candy are Starbursts

# 4. Taking a look at pricepercent

```{r}
# package to avoid over-plotting
library(ggrepel)

ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 6) +
  theme_bw()
```
> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's Miniatures, Starburst, Sour Patch Kids, and Skittles original candy are highly ranked candy that do not cost a lot of money. 

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular? 

The top. 5 most expensive candy types in the dataset are Nik L Nip, Nestle Smarties, Ring pop, Mr Good Bar, and Hershey's Milk Chocolate. The least popular candy is Nik L Nip. 

```{r}
head(arrange(candy, desc(pricepercent), winpercent))
```

# 5. Exploring the correlation structure

Now that we have explored the dataset a little, we will see how the variables interact with one another. 

```{r}
library(corrplot)

cij <- cor(candy)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

The fruity and chocolate variables are the most anti-correlated, which means that not a lot of fruity chocolate candies exist. 

> Q23. Similarly, what two variables are most positively correlated?

Chocolate and winpercent, and chocolate and bar are very positively correlated. 

# 6. Principal Component Analysis

Let’s apply PCA using the 'prcom()' function to our candy dataset remembering to set the **scale=TRUE** argument.

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```
Let's plot our main results as our PCA "score plot"

```{r}
p <- ggplot(pca$x) +
  aes(PC1, PC2, label=rownames(pca$x)) + 
  geom_point(col=my_cols) +
  geom_text_repel(col=my_cols)
p 
```

Combine PCA data and candy data

```{r}
pcandy <- cbind(candy, pca$x[,1:3])
p <- ggplot(pcandy) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(pcandy),
            label=rownames(pcandy)) +
        geom_point(col=my_cols) + 
  labs(title="Halloween Candy PCA", subtitle = "tan: bar; brown: chocolate; pink: fruity; grey: other")
p
```

```{r}
# library(plotly)
# ggplotly(p)
```


```{r}
ggplot(pca$rotation) +
  aes(PC1, reorder(rownames(pca$rotation), PC1)) +
  geom_col()
```

> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, pluribus, and hard are picked up strongly by PC1 in the positive direction. This makes sense to me since a lot of the fruity candy tend to be in a bag with multiple of them and are generally hard candies. The candy that comes to mind are nerds. 